---
title: "Project 2"
author: "Danuta Grabovska"
date: "2025-06-23"
output: html_document
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r}
knitr::opts_chunk$set(dpi=300, fig.width=7, fig.height=4)
library(tidyverse)
library(forecast)   
library(vars)       
library(car)        
library(neuralnet)  
library(ggplot2)
library(lmtest)
library(dlm)
library(readr)      
library(dplyr)
library(zoo)
library(lubridate)   
library(tibble)
library(fGarch)     
library(tseries)    
library(FinTS)
```

# Task 1

We start from loading the data.
```{r}
www1 <- "https://www.mimuw.edu.pl/~noble/courses/TimeSeries/data/aa-rv-20m.txt"
aa <- read_table(www1, col_names = "vol")
```

```{r}
y <- log(aa$vol)
fit_arima <- Arima(y, order = c(0,1,1))
fit_arima
checkresiduals(fit_arima)
```

Now we need to build a local-level model as stated in the task:
$$\begin{array}{lc}
y_t = \mu_t + e_t & e_t \sim N(0, \sigma_e^2) \\
\mu_t = \mu_{t-1} + \eta_t & \eta_t \sim N(0, \sigma_{\eta}^2)
\end{array}$$

```{r}
theme_set(theme_bw())

buildFun <- function(par){
  dlmModPoly(order = 1, dV = exp(par[1]), dW = exp(par[2]))
}

init <- log(var(y) / 10)
fit  <- dlmMLE(y, parm = c(init, init), build = buildFun)
mod  <- buildFun(fit$par)                       

filt <- dlmFilter(y, mod)                    
smooth <- dlmSmooth(filt)                     

sd_eps  <- sqrt(exp(fit$par[1]))
sd_eta  <- sqrt(exp(fit$par[2]))
cat("Standard deviation for epsilon:", round(sd_eps,4), "   Standard deviation for eta", round(sd_eta,4), "\n")

df <- data.frame(
  t        = seq_along(y),
  obs      = y,
  filtered = dropFirst(filt$m),   
  smoothed = dropFirst(smooth$s)
) |>
  tidyr::pivot_longer(-t, names_to = "series", values_to = "value")

ggplot(df, aes(t, value, colour = series)) +
  geom_line() +
  scale_colour_manual(values = c(obs = "black",
                                 filtered = "red",
                                 smoothed = "blue")) +
  labs(title = "Local-level model",
       x = "observation index", y = "log RV", colour = NULL)
```

# Task 2

As always, we start from loading the dataset:
```{r ex2-pfe}
www2 <- "https://www.mimuw.edu.pl/~noble/courses/TimeSeries/data/m-pfesp-ex9003.txt"
pf <- read_table(www2, col_names = c("pf","sp"))
pf_ts <- ts(pf, start = c(1990,1), frequency = 12)
plot(pf_ts)
```

First we need to fit a fixed-coefficient market model to the Pfizer stock returns. Fixed-coefficient model means CAPM with $\alpha$ and $\beta$ being time-invariant, therefore constant. So it is enough to fit a good old linear model.

```{r}
y  <- pf$pf
x  <- pf$sp
lm_capm <- lm(y ~ x)
coeftest(lm_capm)
```

So our $\alpha$ was estimated as 0.0121814, and $\beta$ as 0.8512568. Both coefficients are statistically significant.

Now it's time for a more interesting part, which is fitting a time-varying CAPM to the Pfizer stock return.
Let's recall what is the form of this model:

$$\begin{array}{ll}
r_t = \alpha_t + \beta_t r_{M,t} + e_t & e_t \sim N(0, \sigma_e^2) \\
\alpha_{t+1} = \alpha_t + \eta_t & \eta_t \sim N(0, \sigma_{\eta}^2) \\
\beta_{t+1} = \beta_t + \epsilon_t & \epsilon_t \sim N(0, \sigma_{\epsilon}^2)
\end{array}$$


```{r}
buildCAPM <- function(par){
  dlmModReg(
    X   = x,
    dV  = exp(par[1]),             # observation variance  (scalar)
    dW  = c(exp(par[2]),           # α_t innovation variance
            exp(par[3]))           # β_t innovation variance
  )
}
init <- rep(log(var(y)/20), 3)     # safe starting point
fit  <- dlmMLE(y, parm = init, build = buildCAPM, control = list(maxit = 200))
mod  <- buildCAPM(fit$par)        

# 4. filter & smooth -------------------------------------------------
flt <- dlmFilter(y, mod)
sm  <- dlmSmooth(flt)

alpha_sm <- drop(sm$s[-1, 1])      # first state  = α_t
beta_sm  <- drop(sm$s[-1, 2])      # second state = β_t

plot(beta_sm, type = "l", col = 4,
     ylim = range(c(alpha_sm, beta_sm)),
     ylab = "", xlab = "month",
     main = expression(paste("Smoothed  ", beta[t], " (blue)  &  ", alpha[t], " (red)")))
lines(alpha_sm, col = 2)
legend("topleft",
       legend = c(expression(beta[t]), expression(alpha[t])),
       col = c(4,2), lty = 1, bty = "n")

se_eps  <- sqrt(exp(fit$par[1]))
se_alpha<- sqrt(exp(fit$par[2]))
se_beta <- sqrt(exp(fit$par[3]))
cat("σ_ε =", round(se_eps ,4),
    " σ_α =", round(se_alpha,4),
    " σ_β =", round(se_beta ,4), "\n")


```

# Task 3

As always, we start from loading the dataset:

```{r ex3-ppi}
www3 <- "https://www.mimuw.edu.pl/~noble/courses/TimeSeries/data/m-ppiaco4709.txt"
ppi <- read_table(www3, col_names = c("y","m","d","ppi"))
ppi_ts <- ts(ppi$ppi, start = c(1947,1))
plot(ppi_ts)
```

```{r}
y    <- diff(log(ppi$ppi)) - mean(diff(log(ppi$ppi)))
y_ts <- ts(y)
ggtsdisplay(y_ts)
```

As stated in the task, let's fit a simple AR(3) model.

```{r}

ar3 <- arima(y, order = c(3,0,0))
ar3
coeftest(ar3)
checkresiduals(ar3)
```
Okay, so the model we fitted is $y_t = 0.2873 y_{t-1} + 0.1226 y_{t-2} + 0.0903 y_{t-3}$.

```{r}
Tlen  <- length(y)

build_AR3_me <- function(par){
  phi  <- par[1:3]
  Wvar <- exp(par[4])
  Vvar <- exp(par[5])

  GG   <- rbind(phi, c(1,0,0), c(0,1,0))     
  FF   <- matrix(c(1,0,0), 1, 3)            

  dlm(
    FF = FF,
    V  = matrix(Vvar, 1, 1),                 
    GG = GG,
    W  = matrix(0, 3, 3) + diag(c(Wvar, 0, 0)), 
    m0 = c(0,0,0),
    C0 = diag(1e6, 3)
  )
}


start <- c(0.3, 0.1, 0.05, log(var(y)/50), log(var(y)/10))
fit <- dlmMLE(y, parm = start, build = build_AR3_me, control = list(maxit=200))
mod <- build_AR3_me(fit$par)     

phi_hat  <- round(fit$par[1:3], 4)
sig_eta  <- round(sqrt(exp(fit$par[4])), 5)
sig_eps  <- round(sqrt(exp(fit$par[5])), 5)

cat("Fitted AR(3) coefficients: φ1 =", phi_hat[1],
    ", φ2 =", phi_hat[2], ", φ3 =", phi_hat[3], "\n")
cat("Innovation s.d.  σ_η =", sig_eta,
    ";  measurement-error s.d.  σ_ε =", sig_eps, "\n")

kf  <- dlmFilter(y, mod)
ks  <- dlmSmooth(kf)

x_sm <- drop(ks$s[-1,1])         
res_filt <- residuals(kf, type = "raw")$res

par(mfrow=c(2,1), mar=c(3,4,2,1))
plot(x_sm, type="l", col=4, main="Smoothed estimate of x_t", ylab="x_t")
abline(h=0, col="grey")

plot(res_filt, type="h", col=2, main="Filtered response residuals",
     ylab="e_t  (one-step)", xlab="time")
abline(h=0, col="grey")
```

# Task 4

```{r ex4-ew}
www4 <- "https://www.mimuw.edu.pl/~noble/courses/TimeSeries/data/d-ibm3dxwkdays8008.txt"
ew <- read.table(www4, col.names = c("y","m","d","ibm","vw","ew","sp","mon","tu","we","th","fr"))

ew <- within(ew, {ew  <- as.numeric(ew);
                  mon <- as.numeric(mon); 
                  tu <- as.numeric(tu); 
                  we  <- as.numeric(we);
                  th <- as.numeric(th);
                  fr <- as.numeric(fr)})
ew <- subset(ew, !is.na(ew))           

fit_lm   <- lm(ew ~ mon + tu + we + th + fr, data = ew)
resid_lm <- residuals(fit_lm)

tsdisplay(resid_lm, main = "Regression residuals")
checkresiduals(resid_lm)
```
There definitely are some correlations in the regression residuals - based on ACF, PACF an Ljung-Box test. Therefore, we need to build a suitable time series model on these residuals. Auto-fitted ARIMA still leaves some autocorrelation, therefore we use ARMA_GARCH. The coefficients were chosen manually based on their statistical significance, because (once again) auto-fitted ARMA was quite poor.
```{r}
garch_t <- garchFit(~ arma(3,2) + garch(1,1),
                    data = resid_lm,
                    cond.dist = "std",
                    include.mean = FALSE,
                    trace = FALSE)

summary(garch_t)

par(mfrow = c(2, 2), mar = c(3,4,2,1))

plot(residuals(garch_t), type="l",
     main = "GARCH standardized residuals")

acf(residuals(garch_t), main = "ACF std. resid.")
acf(residuals(garch_t)^2, main = "ACF squared resid.")

archTest <- ArchTest(residuals(garch_t), lags = 12)

mtext(paste("ARCH-LM p-value =",
            round(archTest$p.value, 4)), side = 3, line = -2)


par(mfrow=c(1,1))

sigma_t <- garch_t@sigma.t            
ew$cond_vol <- sigma_t

cat("Last conditional volatility estimate (annualised):",
    round(tail(sigma_t*sqrt(252), 1), 4), "\n")
```

# Task 5

```{r task5-nn, echo=TRUE, message=FALSE, warning=FALSE}
theme_set(theme_bw())

ge <- read_table(
  "https://www.mimuw.edu.pl/~noble/courses/TimeSeries/data/m-ge2608.txt")
r <- as.numeric(ge$rtn)                     
stopifnot(!anyNA(r))

lagmat <- function(x, k) embed(x, k + 1)[ , -1]

train <- head(r, length(r) - 36)
test  <- tail(r, 36)

train_reg <- data.frame(lagmat(train, 3))
train_reg$y <- tail(train, -3)

set.seed(123)
nn_reg <- neuralnet(y ~ ., data = train_reg,
                    hidden = 2, linear.output = TRUE)

X_test <- lagmat(c(tail(train, 3), test), 3)           
pred_reg <- as.numeric(compute(nn_reg, X_test)$net)
mse_reg  <- mean((test - tail(pred_reg, 36))^2)

df_plot <- data.frame(
  idx  = seq_along(test),
  actual = test,
  pred   = tail(pred_reg, 36))

p1 <- ggplot(df_plot, aes(idx)) +
  geom_line(aes(y = actual, colour = "actual")) +
  geom_line(aes(y = pred,   colour = "forecast")) +
  labs(title = "Test-set: GE numeric forecast (3-2-1 NN)",
       y = "return", x = "month") +
  scale_colour_manual(values = c(actual = "black", forecast = "red"),
                      name = "")
p2 <- ggplot(df_plot, aes(idx, actual - pred)) +
  geom_hline(yintercept = 0, colour = "grey") +
  geom_line(colour = "steelblue") +
  labs(title = "Forecast residuals", y = "error")

print(p1); print(p2)
cat("Numeric-NN test-set MSE =", round(mse_reg, 6), "\n")

dr <- as.integer(r > 0)          # 1 = up, 0 = down

train_cls <- data.frame(lagmat(train, 3),          
                        lagmat(dr[1:length(train)], 3))  
names(train_cls) <- paste0("x", 1:6)
train_cls$target <- tail(dr[1:length(train)], -3)

up_idx   <- which(train_cls$target == 1)
down_idx <- which(train_cls$target == 0)

set.seed(7)
dup_down <- sample(down_idx, length(up_idx), replace = TRUE)
balanced <- rbind(train_cls[up_idx, ], train_cls[dup_down, ])

nn_bal <- neuralnet(target ~ ., balanced,
                    hidden = 5, linear.output = FALSE)


dr_test <- as.integer(test > 0)

X_test_cls <- data.frame(
  lagmat(c(tail(train, 3), test), 3),
  lagmat(c(tail(dr[1:length(train)], 3), dr_test), 3))
names(X_test_cls) <- paste0("x", 1:6)
X_test_cls <- X_test_cls[-1, ]          

prob      <- as.numeric(compute(nn_bal, X_test_cls)$net) 
pred_dir  <- as.integer(prob > 0.5)

mse_cls  <- mean((dr_test - pred_dir)^2)
hit_rate <- mean(dr_test == pred_dir)
dr_test <- dr_test[-1] 

cat("Direction-NN test-set MSE =", round(mse_cls, 6),
    " (hit-rate =", round(hit_rate, 3), ")\n")
table(Pred = pred_dir, Actual = dr_test)

```
After balancing the classes and retraining a 6-5-1 feed-forward network, the out-of-sample hit-rate is 0.50, equal to random guessing. This indicates that $\{r_{t−1},r_{t−2},r_{t−3},$signs$\}$ carries no predictive power for next-month GE direction. The numeric 3-2-1 network did reduce MSE in levels, but direction remains unpredictable.

# Task 6

```{r ex6-usdeur, echo=TRUE, message=FALSE, warning=FALSE}
www6 = "https://www.mimuw.edu.pl/~noble/courses/TimeSeries/data/USDEUR.csv"

df_raw <- read_csv(
  www6,
  col_types = cols(.default = col_character())) %>%
  transmute(date  = as.Date(`End Date`, format = "%Y-%m-%d"),
            price = as.numeric(`USD/EUR`)) %>%
  drop_na()

stopifnot(!anyNA(df_raw$date), !anyNA(df_raw$price))
df_raw_ts <- ts(df_raw$price)
ggtsdisplay(df_raw_ts)
```
We decided to select 2011 as the training year and 2012 as testing year. Then, we fit auto-ARIMA on the training year. We decided to stop on ARIMA for several reasons: the simplicity of the model and its performance, as well as the fact that more complex neural or factor-based models add computational burden, but, per recent literature (https://www.sciencedirect.com/science/article/pii/S0164070414000470?utm_source=chatgpt.com, https://rogoff.scholars.harvard.edu/sites/g/files/omnuum5901/files/rogoff/files/51_jie1983.pdf), rarely outperform the random-walk benchmark at the weekly horizon, so ARIMA is the most defensible baseline.
```{r}
year1 <- 2011                      # training year
year2 <- year1 + 1                 # test / forecast year

train <- filter(df_raw, year(date) == year1)
test  <- filter(df_raw, year(date) == year2)

if(nrow(train) == 0 || nrow(test) == 0)
  stop("Selected years not present in the series")

ts_train <- ts(train$price, frequency = 52, start = c(year1, 1))
fit <- auto.arima(ts_train, stepwise = FALSE, approximation = FALSE)

fit
checkresiduals(fit)
coeftest(fit)
```
After that, we're ready to go with one-step forecasting loop:
```{r}
h <- nrow(test) 
fc <- numeric(h)
state <- ts_train

for(i in seq_len(h)){
  fit   <- Arima(state, model = fit)          
  fc[i] <- forecast(fit, h = 1)$mean          
  state <- ts(c(state, test$price[i]),        
              frequency = 52, start = start(state))
}
```

Finally, let's visualise what we got:
```{r}
plot(test$date, test$price, type="l", col="black",
     main="Weekly forecast (red) vs actual (black)")
lines(test$date, fc, col="red")
legend("topleft", c("actual","forecast"), lty=1, col=c("black","red"))
err <- test$price - fc
cat("RMSE =", sqrt(mean(err^2)),
    ";  MAE =", mean(abs(err)), "\n")
plot(test$date, err, type="h", col="steelblue",
     main="One-step forecast errors"); abline(h = 0, col = "grey")
out <- tibble(date = test$date,
              forecast = fc)
out
```
As we can see, the fit is definitely acceptable along with its RMSE and MAE. One-step forecast errors are small and visually resemble white noise.
